---
title: "Preprocess DNA methylation"
author: "Kevin Thomas"
date: "6/20/2023"
output:
  rmarkdown::github_document:
    html_preview: false
    toc: true
    toc_depth: 3
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preprocessing cell type-specific DNA methylation array data

## Setup
```{r, setupPkgs}
BiocManager::install(
  c("minfi", "IlluminaHumanMethylationEPICmanifest", "IlluminaHumanMethylationEPICanno.ilm10b4.hg19"),
  update = FALSE
)
BiocManager::install("preprocessCore", configure.args = c(preprocessCore = "--disable-threading"), force= TRUE, update=FALSE, type = "source")
library(minfi)
library(tidyverse)
```

## Raw data import

```{r, import}
# Read sample sheet
data_dir <- "~/workspace/datasets/methylation/lra-bms-methylation"
sheetEpic <- read.metharray.sheet(base=data_dir, pattern="csv$", ignore.case=T, verbose=T, recursive=T)
# Fix file names
sheetEpic$Basename <- paste0(data_dir, "/", sheetEpic$Array, "/", sheetEpic$Slide)

# Import raw data
rawData.idat <- read.metharray.exp(targets = sheetEpic, recursive = T, verbose = T, force = T)
```

## Quality control
### Beta value distribution
```{r, betaDens}
densityPlot(dat = rawData.idat)
```
### Filter samples by average detection p value
```{r, detP, fig.height=8}
detP <- detectionP(rawData.idat)
meanDetectionPValue <- colMeans(detP)
enframe(meanDetectionPValue, name = "Sentrix_ID", value = "meanDetPVal") %>%
  left_join(sheetEpic, by = c("Sentrix_ID"="Slide")) %>%
  ggplot(aes(x = Sentrix_ID, y = meanDetPVal)) +
  geom_col() +
  geom_hline(yintercept = 0.05, col = "red") +
  facet_wrap(~Plate, ncol = 1, scales = "free_x") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90))
rm(detP)
gc(verbose = FALSE) # memory management
```


Sample with Sentrix_ID 206838570052_R02C01 should be removed.

## Remove sample and re-import data
```{r, remove}
# Edit sample sheet
sheetEpic_filt <- sheetEpic %>%
  filter(Slide != "206838570052_R02C01")

# Import filtered samples only data
rawData.filt.idat <- read.metharray.exp(targets = sheetEpic_filt, recursive = T, verbose = F, force = F) 
```
## Normalization
```{r, norm}
normalized.idat <- preprocessFunnorm(rawData.filt.idat, bgCorr=T, dyeCorr=T, verbose=T)
gc(verbose = FALSE) # For memory management
```

### Probe filtering
```{r, filter}
detP <- detectionP(rawData.filt.idat)
detP <- detP[match(featureNames(normalized.idat), rownames(detP)),]
keepP <- rowSums(detP < 0.05) == ncol(normalized.idat)
normalized.idat <- normalized.idat[keepP,]
```

Removing all CpG probes with a p-value of detection less than 0.05 in all samples.

### Data exploration and outlier removal
``` {r, pca}
# Get M values
methyl_m <- getM(normalized.idat)
gc(verbose = FALSE) # For memory management

# Perform PCA
pca <- prcomp(
  x = t(methyl_m),
  center = TRUE,
  scale. = TRUE
)

# Identify outliers
outliers <- pca$x[,1:2] |>
  bind_cols(sheetEpic_filt) |>
  group_by(group1) |>
  filter(
    abs(PC1 - mean(PC1)) > 3*sd(PC1),
    abs(PC2 - mean(PC2)) > 3*sd(PC2)
  ) |>
  pull(Sample)

# Visualize on PCA plot
pca$x[,1:2] |>
  bind_cols(sheetEpic_filt) |>
  mutate(is_outlier = Sample %in% outliers) |>
  ggplot(aes(x = PC1, y = PC2, col = is_outlier)) +
  geom_text(aes(label = Sample)) +
  coord_fixed() +
  theme_bw()
```
Samples AD058-41 and AD046-31 should be removed as outliers as well.

## Final data assembly
``` {r, final_select}
# Final sample selection
sheetEpic_filt <- sheetEpic %>%
  filter(Slide != "206838570052_R02C01", !Sample %in% outliers)

# Import
rawData.filt.idat <- read.metharray.exp(targets = sheetEpic_filt, recursive = T, verbose = T, force = T)

# Normalize
normalized.idat <- preprocessFunnorm(rawData.filt.idat, bgCorr=T, dyeCorr=T, verbose=T)

# Filter genes
normalized.idat <- normalized.idat[keepP,]

# Extract beta and M-value matrices
methyl_b <- getBeta(normalized.idat)
methyl_m <- getM(normalized.idat)
```

## Save data
```{r, save}
# Save
saveRDS(methyl_b, file = "../data/betaMatrix.RDS")
saveRDS(methyl_m, file = "../data/mValMatrix.RDS")
saveRDS(sheetEpic_filt, file = "../data/methylMetaData.RDS")
```

```{r sessionInfo}
sessionInfo()
```